High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-11-10 doc/llr-pldi2013/interface.tex
some edits and working on the reduction story

\subsection{Barriers}
\label{subsec:barriers}
The common synchronization pattern in which multiple consumers are dependent on
a set of producers is supported by a {\tt Barrier} object (lines 25-29 of Figure~\ref{fig:runtimeapi}), which is similar
to a user event, but does not trigger until the expected number of {\tt arrive}
operations have occurred.  To better support
composability and asynchronous operation, our interface differs from barriers in other programming
models\cite{MPI} in three fundamental ways:
%The common synchronization pattern in which multiple consumers are dependent on a 
%(potentially different) set of producers is supported by a {\tt Barrier} object,
%which is similar to a {\tt UserEvent}, but does not trigger until the expected
%number of {\tt arrive} operations have occurred.  Barrier constructs exist in other
%programming models\cite{MPI}, but in addition to the decoupling of the waiters from
%the arrivers, our barrier interface has two further improvements to better support
%composability and asynchronous operation:
\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
\item The set of producers arriving at the barrier can be different from the set of consumers waiting on the barrier.
\item An arrival operation can be made dependent on another event, allowing a client to
express that an operation should be completed before the barrier can trigger.
\item The expected arrival count of a barrier can be dynamically modified, allowing
producer tasks to be subdivided as necessary, even after the barrier has been created.
\end{itemize}

@@

Events and barriers express ordering properties between operations, but 
%in many cases 
ordering is often too strict a requirement.  For many applications access to data need only be atomic and
not necessarily ordered.  Traditionally locks have been used to serialize access to
data.  However, all lock implementations that we are aware of require either blocking
or spinning, neither of which composes well with asynchronous operations.
{\em Deferred locks} are a new synchronization mechanism that allows for synchronization
in an completely asynchronous environment.

@@

The interface for deferred locks is shown in lines 30-38 of Figure~\ref{fig:runtimeapi}.
Unlike traditional blocking locks, the {\tt lock} method (line 32) doesn't
wait if the lock is in use but instead always returns immediately with an event that will be triggered
when the lock has been acquired.  Like a processor's spawn method, the lock method also 
takes an optional event parameter as a precondition.  Similarly, the {\tt unlock} 
method (line 33) takes an optional event precondition parameter.

@@

When a lock is used to mediate access to a piece of data, the first thing a task will usually
do after being granted the lock is access the data.  To be able to hide the latency of that
access as well, an additional feature of locks in our interface is the association of a lock with a
{\em payload}, a small (i.e. less than 4KB) 
piece of data that is moved around with the lock and guaranteed to be coherent while 
the lock is held.  The size of the payload is specified when the lock is created (line 35 of Figure~\ref{fig:runtimeapi}) and
the pointer to the local copy of the payload can be obtained from the {\tt payload\_ptr} method
(line 36).

@@

\subsection{Physical Regions}
\label{subsec:phyreg}
In distributed machines with discrete memories, data movement often consists of more
than a simple copy operation.  Many applications perform operations and transformations on
their data in conjunction with data movement for higher performance.  One common example is 
that applications accumulate reduction operations in a buffer and then, as part of a copy operation,
the reduction operations are applied to a destination buffer on the receiving side of the copy.  
To support these kinds of conjoined operations in an asynchronous environment, our interface 
must be aware of the structure of data so it can perform accompanying operations with 
data movement operations.  To describe the structure of data our interface uses {\em physical regions.}

>>>>

\subsection{Barriers}
\label{subsec:barriers}
The common synchronization pattern in which multiple consumers are dependent on
a set of producers is supported by a {\tt Barrier} object (lines 25-29 of Figure~\ref{fig:runtimeapi}), which is similar
to a user event, but does not trigger until the expected number of {\tt arrive}
operations have occurred.  To better support
composability and asynchronous operation, our interface differs from barriers in other programming
models\cite{MPI} in three fundamental ways:
%The common synchronization pattern in which multiple consumers are dependent on a 
%(potentially different) set of producers is supported by a {\tt Barrier} object,
%which is similar to a {\tt UserEvent}, but does not trigger until the expected
%number of {\tt arrive} operations have occurred.  Barrier constructs exist in other
%programming models\cite{MPI}, but in addition to the decoupling of the waiters from
%the arrivers, our barrier interface has two further improvements to better support
%composability and asynchronous operation:
\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
\item The set of producers arriving at the barrier can be different from the set of consumers waiting on the barrier.
\item An arrival operation can be made dependent on another event, allowing a client to
express that an operation should be completed before the barrier can trigger.
\item The expected arrival count of a barrier can be dynamically modified, which we
have found useful in supporting nested parallelism in higher-level languages.
%allowing
%producer tasks to be subdivided as necessary, even after the barrier has been created.
\end{itemize}

@@

Events and barriers express ordering properties between operations, but 
%in many cases 
ordering is often too strict a requirement.  For many applications access to data need only be atomic and
not necessarily ordered.  Traditionally locks have been used to serialize access to
data.  However, all lock implementations that we are aware of require either blocking
or spinning, neither of which composes well with asynchronous operations.
{\em Deferred locks} are a new synchronization mechanism that allows for synchronization
without blocking or spinning.
%in an completely non-blocking environment.

@@

The interface for deferred locks is shown in lines 30-38 of Figure~\ref{fig:runtimeapi}.
Unlike traditional blocking locks, the {\tt lock} method (line 32) doesn't
wait if the lock is in use but instead always returns immediately with an event that will be triggered
when the lock has been acquired.  Like a processor's spawn method, both the {\tt lock} 
and {\tt unlock} methods take an optional event paramenter as a precondition (lines 32-33).

@@

When a lock is used to mediate access to a piece of data, the first thing a task will usually
do after being granted the lock is access the data.  To hide the latency of that
access as well, an additional feature of locks in our interface is the association of a lock with a
{\em payload}, a small (i.e. less than 4KB) 
piece of data that is moved around with the lock and guaranteed to be coherent while 
the lock is held.  The size of the payload is specified when the lock is created (line 35 of Figure~\ref{fig:runtimeapi}) and
the pointer to the local copy of the payload can be obtained from the {\tt payload\_ptr} method
(line 36).

@@

\subsection{Physical Regions}
\label{subsec:phyreg}
%In distributed machines with discrete memories, data movement often consists of more
%than a simple copy operation.  Many applications perform operations and transformations on
%their data in conjunction with data movement for higher performance.  One common example is 
%that applications accumulate reduction operations in a buffer and then, as part of a copy operation,
%the reduction operations are applied to a destination buffer on the receiving side of the copy.  
%To support these kinds of conjoined operations in an asynchronous environment, our interface 
%must be aware of the structure of data so it can perform accompanying operations with 
%data movement operations.  To describe the structure of data our interface uses {\em physical regions.}
Current APIs for moving data in machines with discrete memories only support untyped
copies of a buffer of data.  This restriction requires programmers to manage both the layout of
data and the marshalling of data for communication.  {\em Physical regions} are a
mechanism for decoupling the placement of data from its layout.  Physical regions make
the runtime aware of the structure of elements in the allocation and thereby enable multiple data
layouts, such as specialized layouts for buffering individual reduction operations.

### 2012-11-10 doc/llr-pldi2013/intro.tex
some edits and working on the reduction story

Most current programming systems for large-scale parallel processing provide some variant
of the {\em bulk-synchronous} execution model\cite{Valiant90}.  Bulk-synchronous models
partition computations into {\em phases} that are either parallel computation, data
communication, or synchronization.  At any given point in time all processors
are engaged in the same kind of phase.  Bulk-synchrony
makes it easy to reason about parallel execution but
hinders attempts to hide the increasingly long-latency communication
or memory movement phases.  
%In older machines this was not a problem because slower
%processors and smaller machines meant computation phases were much larger than
%communication or synchronization phases.  However, today's machines are much
%larger with faster processors, causing an increasing percentage of time
%to be spent in the communication and synchronization phases.

@@

We propose a new low-level interface for programming modern
supercomputers that is fully asynchronous. In our system, one can
asynchronously compose computation, data movement, and
synchronization.  The basic mechanism that enables composition of 
asynchronous operations
is an {\em event} primitive.  Events provide a mechanism
for naming a point in the future when an asynchronous operation
(computation, data copy, or synchronization) will complete.  Every
call to perform an operation $p$ in our runtime returns immediately
with an event that {\em triggers} when $p$ has completed.
Furthermore, every call to perform an operation $p$ can take as a
precondition an event that must trigger before $p$ can begin.
%Any number of operations can be depedendent on a single event, and
%a events can be merged to allow an operation to wait for multiple
%events before it begins.
Events can be merged into new events and multiple operations
can use an event as a precondition.
  Using
events clients can compose arbitrary chains of dependent
operations, which the runtime is free to execute in any way that
respects the event dependences.  With this freedom the runtime can
schedule operations to optimize throughput while hiding the long
latency operations.

@@

\item We describe a physical region system that provides structural information about
data layout.  The structural information provided by physical regions enables an extensible
interface for conjoining operations on data (e.g. reductions) with asynchronous data movement for 
further latency hiding (Section~\ref{subsec:phyreg}).

\item Our approach depends on the efficient handling of very large numbers of events.
We describe a distributed implementation of our interface that requires no global coordination
and has modest local storage costs.  We describe optimizations for the implementation of events, 
deferred locks, and reductions (Section~\ref{sec:impl}).

\item We report on micro-benchmarks that are designed to stress
our implementation.  The results indicate that our implementation approaches the performance of
the underlying hardware (Section~\ref{sec:micro}).

\item We demonstrate that our interface is powerful enough to support Legion \cite{Legion12}, a higher-level
programming system, and that real Legion applications programmed to our interface
are capable of achieving high-performance (Section~\ref{sec:apps}).  Furthermore, we demonstrate
that the same applications written in a bulk-synchronous model incur performance penalties between 22-135\%.
\end{itemize}

>>>>

Most current programming systems for large-scale parallel processing provide some variant
of the {\em bulk-synchronous} execution model\cite{Valiant90}.  Bulk-synchronous models
partition computations into {\em phases} that are either parallel computation, data
communication, or synchronization.  At any given point in time all processors
are engaged in the same kind of phase.  Bulk-synchrony
makes it easy to reason about parallel execution but
hinders attempts to hide the increasingly long-latency communication
or synchronization phases.  
%In older machines this was not a problem because slower
%processors and smaller machines meant computation phases were much larger than
%communication or synchronization phases.  However, today's machines are much
%larger with faster processors, causing an increasing percentage of time
%to be spent in the communication and synchronization phases.

@@

We propose a new low-level interface for programming modern
supercomputers that is fully asynchronous. In our system, one can
asynchronously compose computation, data movement, and
synchronization.  The basic mechanism that enables composition of 
asynchronous operations
is an {\em event} primitive.  Events provide a mechanism
for naming a point in the future when an asynchronous operation
(computation, data copy, or synchronization) will complete.  Every
call to perform an operation $p$ in our runtime returns immediately
with an event that {\em triggers} when $p$ has completed.
Furthermore, every call to perform an operation $p$ can take as a
precondition an event that must trigger before $p$ can begin.
%Any number of operations can be depedendent on a single event, and
%a events can be merged to allow an operation to wait for multiple
%events before it begins.
Events can be merged into new events and multiple operations
can use an event as a precondition.
  Using
events clients can compose arbitrary chains of dependent
operations, which the runtime is free to execute in any way that
respects the event dependences.  With this freedom the runtime can
schedule operations to optimize throughput while hiding long
latency operations.

@@

%\item We describe a physical region system that provides structural information about
%data layout.  The structural information provided by physical regions enables an extensible
%interface for conjoining operations on data (e.g. reductions) with asynchronous data movement for 
%further latency hiding (Section~\ref{subsec:phyreg}).
\item We describe a physical region system that encourages efficient movement of data
via bulk transfers.  The structural information provided by physical regions also enables
special reduction instances that buffer individual reduction operations for bulk
transfer and application (Section~\ref{subsec:phyreg}).

\item Our approach depends on the efficient handling of very large numbers of events.
We describe a distributed implementation of our interface that requires no global coordination
and has modest local storage costs.  We describe an optimized implementation of events, 
deferred locks, and reductions (Section~\ref{sec:impl}).

\item We report on micro-benchmarks that stress
our implementation.  The results indicate that our implementation approaches the performance of
the underlying hardware (Section~\ref{sec:micro}).

\item We demonstrate that our interface is powerful enough to support Legion \cite{Legion12}, a higher-level
programming system, and that real Legion applications programmed to our interface
are capable of achieving high-performance (Section~\ref{sec:apps}).  Furthermore, we demonstrate
that the same applications written in a bulk-synchronous model incur slowdowns between 22-135\%.
\end{itemize}

### 2012-11-10 doc/llr-pldi2013/legion-llr.tex
First commit on new low-level runtime paper for pldi
redid reduction stuff, monkeyed with figures to put examples with event graphs
some edits and working on the reduction story

\documentclass{sig-alternate}

\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{listings}
\usepackage{tikz}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{cancel}
\usepackage{graphics}
\usepackage{url}

\newcommand{\infrule}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
\newcommand{\deref}{\ast}
\newcommand{\rread}[1]{\mbox{\em Read}(#1)}
\newcommand{\rwrite}[1]{\mbox{\em Write}(#1)}
\newcommand{\lca}[2]{#1 \sqcup #2}
\newcommand{\rleq}{\leq}
\newcommand{\interval}[1]{\mbox{\em interval}(#1)}
\newcommand{\context}[1]{\mbox{\em context}(#1)}
\newtheorem{theorem}{Theorem} 
\newtheorem{lemma}[theorem]{Lemma}

@@

@@

\title{High-Performance Language Primitives for Distributed Architectures using Asynchronous Events}
% Commented out for double-blind goodness
%% \numberofauthors{3}
%% \author{
%% \alignauthor Sean Treichler\\
%%              \affaddr{Stanford University} \\
%%              \email{sjt@cs.stanford.edu}
%% \alignauthor Michael Bauer \\
%%              \affaddr{Stanford University} \\
%%              \email{mebauer@cs.stanford.edu}
%% \alignauthor Alex Aiken \\
%%              \affaddr{Stanford University} \\
%%              \email{aiken@cs.stanford.edu}
%% }
\maketitle

\begin{abstract}
%The limit is 12 pages, including bibliography and appendices.
%Deadlines:
%\begin{tabular}{ll}
%Abstract deadline & July 16, 2012 3pm PDT \\
%Paper deadline & July 23, 2012 3pm PDT \\
%Author response & around October 15, 2012 \\
%Author notification & November 9, 2012
%\end{tabular}
We present a low-level runtime interface for current highly distributed, heterogeneous
parallel machines that provides a fully asynchronous 
programming model, allowing all types of runtime actions, including synchronization,
to be non-blocking.  Our interface includes mechanisms for the creation and movement
of data through the memory hierarchy as well as {\em deferred locks}, a novel synchronization primitive.
Asynchrony is exposed by the interface using a distributed event system.  All
operations return an event to be triggered when the operation completes.  Events
can be chained together and used to capture dependences between operations.

@@

{
\small
\bibliographystyle{abbrv}
\bibliography{bibliography}
}

>>>>

\documentclass[9pt,nocopyrightspace,preprint]{sigplanconf}
\bibliographystyle{plainnat}
\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{listings}
\usepackage{tikz}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{cancel}
\usepackage{graphics}
\usepackage{url}

\newcommand{\infrule}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
\newcommand{\deref}{\ast}
\newcommand{\rread}[1]{\mbox{\em Read}(#1)}
\newcommand{\rwrite}[1]{\mbox{\em Write}(#1)}
\newcommand{\lca}[2]{#1 \sqcup #2}
\newcommand{\rleq}{\leq}
\newcommand{\interval}[1]{\mbox{\em interval}(#1)}
\newcommand{\context}[1]{\mbox{\em context}(#1)}
\newtheorem{theorem}{Theorem} 
\newtheorem{lemma}[theorem]{Lemma}
\newcommand{\texcomment}[1]{}

@@

\newsavebox{\mylistingbox}

@@

\title{High-Performance Language Primitives for Distributed Architectures using Asynchronous Events}
% Commented out for double-blind goodness
%% \numberofauthors{3}
%% \author{
%% \alignauthor Sean Treichler\\
%%              \affaddr{Stanford University} \\
%%              \email{sjt@cs.stanford.edu}
%% \alignauthor Michael Bauer \\
%%              \affaddr{Stanford University} \\
%%              \email{mebauer@cs.stanford.edu}
%% \alignauthor Alex Aiken \\
%%              \affaddr{Stanford University} \\
%%              \email{aiken@cs.stanford.edu}
%% }
\authorinfo{}{}{}
\maketitle

\begin{abstract}
%The limit is 12 pages, including bibliography and appendices.
%Deadlines:
%\begin{tabular}{ll}
%Abstract deadline & July 16, 2012 3pm PDT \\
%Paper deadline & July 23, 2012 3pm PDT \\
%Author response & around October 15, 2012 \\
%Author notification & November 9, 2012
%\end{tabular}
We present a low-level runtime interface for current highly distributed, heterogeneous
parallel machines that provides a fully asynchronous 
programming model, allowing all runtime actions, including synchronization,
to be non-blocking.  Our interface includes mechanisms for the creation and movement
of data through the memory hierarchy as well as {\em deferred locks}, a novel synchronization primitive.
Asynchrony is exposed by the interface via a distributed event system.  All
operations return an event that will be triggered when the operation completes.  Events
can be chained together and used to capture dependences between operations.

@@

{
\bibliography{bibliography}
}
